---
title: "Team"
author: "Kellan Bouwman & Felix Baez-Santiago"
date: "10/11/2020"
output: html_document
---

## Part one: Cleaning data
Although the three Excel sheets are loaded into R, they are not useful before some cleaning. Manipulate the three imported data frames, so that in the cleaned dataset:

1. The player names and opponent names are factors.

     Answer
     
```{r}
# dplyr
library(dplyr)
# ggplot2
library(ggplot2)
#Libary
library(tidyr)
#Libary
library(readxl)
defense <- read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
offensive <- read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
Biography <- read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
```
     

2. All offensive and defensive statistics are numerical. The Weight column in biogrpahy is also numeric. (Optional: Learn to mutate multiple columns systematically using the ?across function in `dplyr)

     Answer
     
```{r}
cleanDef <- defense %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
str(cleanDef)

cleanOff <- offensive %>%
              mutate(across(c("Name", "Opponent_Opponent"), as.factor)) %>%
              mutate_if(is.character, as.numeric, na.rm = TRUE)
cleanOff <- cleanOff %>% 
                replace(is.na(.), 0)
str(cleanOff)
```
     
3. Change the Height column in biogrpahy into numeric. (Hint: Use a unit that works well)

     Answer
     
```{r}
cleanBio <- Biography %>%
              mutate(Name = as.factor(Name)) %>% 
              mutate(Height = gsub("-", ".", Height)) %>% #remove the - from height and convert to feet
              mutate(across(c("Height"), as.numeric)) #turning height into a numeric column
str(cleanBio)
```

By cleaning the three data frames in R, at the end of this step you should have the following three clean data frames: defClean, offClean, and bioClean. Print the structures of these data frames.

## Part two: Tidying
For the following questions, work on the cleaned data frames defClean, offClean, and bioClean.

1. Reformat the defClean data frame into a tidy format using pivot_longer. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named stat.

     Answer
     
```{r}

```
     
2. Compare the distributions of the defensive statistics. What defensive skills are rare?

     Answer
     
```{r}

```
     
3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.

     Answer
     
```{r}

```
     
Now turn our attention to the player biography.

4. Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)

     Answer
     
```{r}

```
     
5. How many players are there from each state?

     Answer
     
```{r}

```
     
## Part three: Joining data frames

1. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.

     Answer
     
```{r}

```
     
2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

     Answer
     
```{r}

```
     
3. Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data here.

     Answer
     
```{r}

```
     